Supporting Information for\\ ``A weakly structured stem for human origins in Africa''

### 2022-07-22 supplement.tex
more supp results

We performed an initial round of optimization including all six populations
with a family of models including single- and multiple-stem scenarios as
described in the previous section. We identified some parameters that reached
consistent values across all models. :
\begin{itemize}
    \item Neanderthal/human split: 550ka
    \item Neanderthal introgression to Eurasian branch: 45ka
    \item Eurasian back-to-African migration: 12ka
\end{itemize}
When testing multiple variations of the more complex models, we kept these
values fixed (see Table~\ref{tab:fixed-params} for a complete list of all fixed
parameters). This reduced the potential for overfitting the more complex
models, while reducing the computational cost of optimization.

Our models also included recent events to account for known migrations,
admixtures, and growths and declines in effective population sizes. Many of
these parameters were fixed based on previous historical, genetic, or
anthropological research, namely
\begin{itemize}
    \item Mende population expansion and Gumuz population decline: 5ka \citep{Gopalan2022-pw}
    \item East African pastoralist to South African admixture: 2ka \citep{Uren2016-nn}
    \item European to South African admixture: 10 generations ago
    \item South African population decline following colonial admixture: 9 generations ago
\end{itemize}
allowed to vary in the fits. The total number of parameters that were
ultimately inferred were 21-31, depending on the complexity
of the model.

In our preliminary models, the unconstrained inferred split times between East African populations 
and Europeans was inferred to have occurred around 40kya \sgcomment{check}, which is 
more recent than we may have expected given 
the well-documented presence of modern humans out of Africa at that time\sgcomment{cite}, as 
well as previous genetics-based dating of the split \sgcomment{cite}. 
The inferred split between East and West African populations occurred about $10ka$ earlier. 
We see two non-exclusive explanations for this apparent discrepancy. 
First, there is uncertainty in our model fit of this parameter.
The $95\%$ confidence interval allows for a split as far back as \sgcomment{$XX$}. 
Second, our confidence intervals do not capture all sources of uncertainty. 
For example, model mis-specification can lead to a mismatch between an inferred split time and the first time populations actually split. 
Our model supposes a single OOA event with constant symmetric migrations 
between populations, which is certainly an oversimplification \cite{}. 
Rather than adding the many parameters required to account for these effects, 
we fixed the split time between Europeans and the East African ancestral population at $50ka,$ and
the split between East/West African population at $60ka$. 
We discuss the sensitivity of our model to these assumptions in Section \ref{sec:change_dates}.

@@

\subsection{Relaxing fixed divergence times}

\begin{itemize}
    \item Fixed to older dates (100ka: E/W split, 60ka: OOA event). These fixed
        values stretched the other inferred parameters to older times, and resulted
        in a worse log-likelihood
    \item Allowed to vary, the best-fit inferred E/W split time and OOA event
        were more recent 35-50ka, instead of our fixed 50-60ka. \aprcomment{I
        still don't find this entirely unreasonable, as E Africans and Eurasian
        populations may be more closely related than the oldest evidence of H.
        sapiens in Europe (up to 50ka), or those earliest H. sapiens are likely
        not to be the direct ancestors of present-day Europeans. Still, it's
        probably too recent of a split, and fixing to 50ka is defensible.}
\end{itemize}

@@

\begin{figure}[ht]
    \centering
    \includegraphics[width=\textwidth]{figures/supp-IM-misspecification}
    \caption{
        \textbf{IM model split times reinferred from simulated data of full
        model.} We simulated 10 diploid individuals under the four highlighted
        models (Sec.~\ref{sec:IM-reinfer}). Using the joint-SFS, we reinferred
        a simpler isolation-with-migration (IM) model for pairs of populations
        including the Nama to explore the effect of model misspecification on
        the earliest inferred divergence times between Khoe-San and other
        populations. We simulated data with Nama individuals sampled at present
        ($t=0$) as well as Nama individuals sampled 2,500 years ago, prior to
        the recent gene flow from East Africans and Europeans. We fit two
        models for each model and population pair: one that allowed for
        migration between the populations (IM), and one that disallowed
        migration (clean split). Despite all original models having recent
        population divergences $\sim120$ka, the reinferred IM models all
        settled on divergence times greater than 200ka and sometimes much
        greater. If the true historical model includes reticulation or
        long-lasting structure, simpler IM models may be severely biased in
        favor of more ancient divergence times.
    }
    \label{fig:supp-IM-misspecification}
\end{figure}

>>>>

We performed an initial round of optimization including all six populations
with a family of models including single- and multiple-stem scenarios as
described in the previous section. We identified some parameters that reached
consistent values across all models:
\begin{itemize}
    \item Neanderthal/human split: 550ka
    \item Eurasian back-to-African migration: 12ka
\end{itemize}
When testing multiple variations of the more complex models, we kept these
values fixed (see Table~\ref{tab:fixed-params} for a complete list of all fixed
parameters). This reduced the potential for overfitting the more complex
models, while reducing the computational cost of optimization.

Our models also included recent events to account for known migrations,
admixtures, and growths and declines in effective population sizes. Many of
these parameters were fixed based on previous historical, genetic, or
anthropological research, namely
\begin{itemize}
    \item Neanderthal introgression to Eurasian branch: 45ka
    \item Mende population expansion and Gumuz population decline: 5ka \citep{Gopalan2022-pw}
    \item East African pastoralist to South African admixture: 2ka \citep{Uren2016-nn}
    \item European to South African admixture: 10 generations ago
    \item South African population decline following colonial admixture: 9 generations ago
\end{itemize}
The total number of parameters that were
ultimately inferred were 21-31, depending on the complexity
of the model.

In our preliminary models, the unconstrained inferred split times between East
African populations and Europeans was inferred to have occurred around $40$ka,
which is more recent than we may have expected given the well-documented
presence of modern humans outside of Africa at that time \citep{}, as well as
previous genetics-based dating of splits between African and Eurasian
populations \citep{}. The unconstrained inferred split between East and West
African populations occurred about $10$ka earlier
(Section~\ref{sec:supp-results}).
%We see two non-exclusive explanations for this apparent discrepancy. First,
%there is uncertainty in our model fit of this parameter. The $95\%$ confidence
%interval allows for a split as far back as \sgcomment{$XX$}.
Across all models, the $95\%$ confidence intervals for the OOA event was
roughly $38 \pm 5$ka, and for the East/West African split was $45 \pm 10$ka.
However, our confidence intervals do not capture all sources of uncertainty.
For example, model mis-specification can lead to a mismatch between an inferred
split time and the first time populations actually split. Our model supposes a
single OOA event followed by constant symmetric migrations between populations,
which is certainly an oversimplification \citep{}. Additionally, most previous
estimates of the time of the OOA expansion compared Eurasian and West African
populations. Our inferred time of the East/West African divergence aligns with
previous estimates of the OOA event at $\approx 60$ka (themselves inferred from
oversimplified demographic models). Rather than adding the many parameters
required to account for these effects, we fixed the split time between
Europeans and the East African ancestral population at $50$ka, and the split
between East/West African population at $60$ka.
%We discuss the sensitivity of our model to these assumptions in Section
%\ref{sec:change-dates}.

@@

\subsection{Relaxed inference of divergence times}
\label{sec:change-dates}

In the primary results presented in the main text
(Tables~\ref{tab:supp-single-origin}--\ref{tab:supp-merger-with-stem-migration}),
we fixed a number of parameters that have been previously inferred or are well
constrained by historical records (Section~\ref{sec:models}), including the
Eurasian out-of-Africa expansion at 50ka and the divergence of Western and
Eastern African groups at 60ka. To evaluate the effects of these choices on the
inference of other parameters in the model, we first fixed those modeled
divergence times to 60ka (for the out-of-Africa expansion) and 100ka (for the
E/W African divergence). In the next section, we allow those parameters to be
fit along with all other model parameters.

By fixing those divergence times to older dates, the inferred models preferred
older dates for the other free divergence time parameters. These dates were
stretched by a similar proportion that our fixed parameters were shifted, so
that the Nama divergence time was inferred to be $\approx150$ka, rather than
$\approx130$ka. Notably, older divergence times resulted in models with worse
likelihoods than our primary presented results. Overall, these inferred models
have a similar interpretation to the primary inference results, and the deep
population structure was unaffected by these choices.

When we allowed the out-of-Africa expansion and Eastern/Western African
divergence times to be jointly fit in model optimization, we found that the
best-fit inferred E/W divergence time and the OOA expansion were more recent
(35--40ka, 45--55ka, resp.). The remainder of the parameters were largely
unaffected by these shifts to more recent divergence times, and our
interpretation of the inferred early history remains unchanged. The $95\%$
confidence intervals around these parameters were relatively narrow ($\approx
5$ka for the OOA split, $\approx 10$ka for the East/West African split),
although this uncertainty does not capture uncertainty due to model
mis-specification (as discussed in Section~\ref{sec:models}).

@@

\begin{figure}[ht]
    \centering
    \includegraphics[width=\textwidth]{figures/supp-IM-misspecification}
    \caption{
        \textbf{IM model split times reinferred from simulated data of full
        model.} We simulated 10 diploid individuals under the four highlighted
        models (Sec.~\ref{sec:IM-reinfer}). Using the joint-SFS, we reinferred
        a simpler isolation-with-migration (IM) model for pairs of populations
        including the Nama to explore the effect of model mis-specification on
        the earliest inferred divergence times between Khoe-San and other
        populations. We simulated data with Nama individuals sampled at present
        ($t=0$) as well as Nama individuals sampled 2,500 years ago, prior to
        the recent gene flow from East Africans and Europeans. We fit two
        models for each model and population pair: one that allowed for
        migration between the populations (IM), and one that disallowed
        migration (clean split). Despite all original models having recent
        population divergences $\sim120$ka, the reinferred IM models all
        settled on divergence times greater than 200ka and sometimes much
        greater. If the true historical model includes reticulation or
        long-lasting structure, simpler IM models may be severely biased in
        favor of more ancient divergence times.
    }
    \label{fig:supp-IM-misspecification}
\end{figure}
